Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-06-05 paper.Rnw
lotta changes to mediation sec 4, #7

The following section will provide background for the case-study,
describing the CTA1 curriculum, the RAND study, and our focus on
hints.
The following three sections will present an observational study,
which discards the control group altogether and estimates the effect
of hint requests on posttest scores;
causal mediation, which decomposes the overall treatment effect into a
component driven by hint requests and another driven by other
mechanisms; and principal stratification, which estimates separate
treatment effects for groups of students who would request hints
differently, if assigned to treatment.
As it turns out, these analyses appear to give contradictory
results---the observational study and mediation analysis estimate a
negative effect of requesting hints, while the principal
stratification analysis finds that students who request more hints
experience larger overall treatment effects.
Section \ref{sec:synthesis} will contrast these findings (including a
possible reconciliation), compare the three methods generally,
and discuss how they may be applied to data from other educational
technology RCTs.
Section \ref{sec:conclusion} will conclude.

@@

\subsection{Estimating the Effects of $H$ on Posttest Scores}

@@

\section{Mediation Analysis}\label{sec:mediation}
The goal of causal mediation analysis is to decompose a treatment
effect into an ``indirect'' or ``mediated''
effect and a ``direct'' effect.
The indirect effect captures the component of the effect that is due
to the mediator: the treatment affected the outcome by affecting the
mediator, which itself, in turn, affected the outcome.
Assignment to the treatment condition allowed students to request hints during their work---how did the availability of hints affect their posttest scores?
The direct effect is the component of the overall effect that operates via other mechanisms.
For instance, how would assignment to CTA1 affect posttest scores were hints not available (or, at least, limited)?

@@

The following subsection reviews the formal definitions of direct and indirect effects.
The next two subsections illustrate the estimation of direct and indirect effects, respectively, in our study of hints.
It turns out that indirect effects are are nearly unidentifiable, without some acrobatics; however, under the identification scheme we present here, estimating indirect effects is exceedingly simple.
Therefore, Section \ref{sec:indirectEst} will mostly consist of a discussion of identification.
Section \ref{sec:directEst} builds on the framework of the previous subsections to estimate direct effects with a regression model.

\subsection{Review: Defining Direct and Indirect Effect}
More precisely, express a subject's potential outcomes as a function
of two variables, treatment assignment $Z$ and hint requests $H$:
$Y(Z,H)$.
Now, if $Z$ and $H$ are both binary, as in our example, subjects
each have four potential outcomes:
\begin{equation*}
Y(0,0);\;Y(1,0);\;Y(0,1);\;Y(1,1)
\end{equation*}
representing the outcomes they would exhibit were they assigned to
control and requested few hints, assigned to
treatment and requested few hints, assigned to control and requested
many hints, or assigned to treatment and requested many hints, respectively.
In the CTA1 trial, however, the potential outcome $Y(0,1)$ is
meaningless: students in the control group had no access to CTA1, and
therefore could not request hints.
This will be an important factor going forward.

@@

Combining potential values for $Y$ and for $H$ yields the fundamental
building blocks of causal mediation analysis
\citep[e.g.][]{vanderweele2015explanation,sales2017mediation}.
Let $Y(z,H(z'))$
represent the outcome that a subject would express given treatment
assignment $z$, but the hint behavior that he or she would have
expressed under assignment $z'$.
When $z=z'$, the original two potential outcomes for $Y$ emerge:
$Y(1,H(1))=Y(Z=1)$ is the outcome exhibited under the treatment, when $H$
takes its potential treatment value; similarly, $Y(0,H(0))=Y(Z=0)$. On
the other hand, $Y(1,H(0))$ and $Y(0,H(1))$ are strictly
counterfactual. $Y(1,H(0))$ gives the outcome value
that would result for a subject assigned to the treatment condition
but who nevertheless requested hints as he or she would have under
control. $Y(0,H(1))$ gives the potential outcome for a subject
assigned to control but who nevertheless requested hints as he or she
would have under treatment.
This last quantity is problematic in our case, since $H(1)$ can take the value 1,
and as we have seen $Y(0,1)$ is undefined.

This framework facilitates precise definitions of direct and indirect
effects; in fact, there are two versions of each \cite[e.g.][]{imai2011unpacking}:
\begin{align*}
\delta(1)&=Y(1,H(1))-Y(1,H(0))\\
\delta(0)&=Y(0,H(1))-Y(0,H(0))\\
\end{align*}
are the indirect effects, contrasting potential outcomes when $H$
varies as it would with varying treatment assignment, but holding the
assignment itself constant at either 1 or 0.
\begin{align*}
\xi(1)&=Y(1,H(1))-Y(0,H(1))\\
\xi(0)&=Y(1,H(0))-Y(0,H(0))\\
\end{align*}
represent the direct effects: holding the value of $H$ constant,
either at its potential value under treatment or control, while
varying treatment assignment.
The total treatment effect, $Y(1,H(1))-Y(0,H(0))$, can then be
decomposed in two ways: as $\delta(1)+\xi(0)$ or as
$\delta(0)+\xi(1)$.
These decompositions can potentially reveal the role hints play in the
CTA1 treatment effect: $\delta(1)$ gives the extent to which
assignment to CTA1 affects a student's posttest score by (possibly)
causing him or her to request hints.
$\xi(0)$ gives the effect of assignment to CTA1 if hints were,
perhaps counterfactually, held at their control level, $H=0$.

\subsection{Identification and Estimation Log Data  Indirect
  Effects}\label{sec:indirectEst}

@@

Dichotomizing hint requests into $H$ provides a
way forward.
The experimental design ensures that $H(0)=0$ for all experimental
participants, since
CTA1 software was unavailable to the control group.
Therefore, $Y(1,H(0))\equiv Y(1,0)$.
Specifically, for treated subjects $i$ for whom $H_i=0$, then, treatment assignment
did not affect $H$, so $H_i(0)=H_i(1)=0$.
%In other words, for these subjects,
%\begin{equation*}
%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
%\end{equation*}
In other words, when $Z_i=1$ and $H_i=0$,
$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
indirect effect.

Of course, $H=0$ in the treatment group typically means requesting
\emph{some} hints, but not many, whereas subjects in the control group
request exactly zero hints.
One way to make sense of mediational estimands involving $H$ is to
imagine an experiment in which students are assigned to different
values of $H$.
\Sexpr{round((1-mixProp)*100)}\% of students in the treatment group
are randomly assigned to $H=1$ and \Sexpr{round((mixProp)*100)}\% to
$H=0$.
This assignment constrains $\bar{h}$, the proportion of problems on which they
may request hints, so that if $H_i=0$,
$\bar{h}_i$ is constrained to be less than
\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))},
but if $H_i=1$ $\bar{h}_i$ must be greater than or equal to
\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
In contrast, all students in the control group are assigned $H=0$ and $\bar{h}=0$.

In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
individual level, as $\delta_i(1)$ and $\xi_i(0)$.
However, individual effects are
not identified without strong assumptions, so the target of
estimation is instead taken to be super-population average effects,
$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$.
Our approach to estimating these
is to estimate the means of each of the component potential outcomes,
$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
Y(0,H(0))=\EE Y(Z=0)$ are non-parametrically identified due to the
randomization of treatment assignment.
Additionally, as described above, $\delta_i(1)=0$ as long as $Z_i=1$ and $H_i=0$.

Treated subjects with $H_i=1$ present more of a challenge.
To estimate indirect effects for these subjects, we may rely on the
same causal setup as in the observational approach:
matched ignorability and the matching design of Section
\ref{sec:observational}.
Each member of the treatment group with $H=1$ is matched to other
treatment-group subjects with $H=0$.
Matched ignorability states that within these matched sets, $Y(1,0)$
is independent of $H$.
Therefore, if subject $i$ is matched to subject $j$, with $H_i=1$, and
$H_j=0$, we may estimate $Y_i(1,0)$ based on the observed
$Y_j=Y_j(1,0)$.
When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
$Y_i(1,0)$ is estimated with the average of their observed outcomes.

Specifically, let
\begin{equation}\label{eq:estY10}
\hat{Y}_i(1,0)=\begin{cases}
 Y_i & H_i=0\\[3ex]
\frac{\displaystyle\sum_{j:M_j=M_i}
  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
\end{cases}
\end{equation}
For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
average outcome of their matched pairs with $H=0$.

Now, for each member $i$ of the treatment group we have observed
$Y_i(1,H(1))=Y_i$, and estimated or inferred
$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
Then, we can estimate $\Delta$, the average indirect effect in the
treatment group, by averaging over the treatment group:
\begin{equation*}
 \hat{\Delta}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
\end{equation*}

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the average difference between the realized outcomes for students
with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)-Y(1,0)|H=1]
\end{equation*}

$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1,H(1))-Y(1,0)]
\end{equation*}
This can be re-written:
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
&\qquad +\EE[Y(1,0)-Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
&=Pr(H=1)TOT
\end{align*}
Since, by definition, $Pr(H=1)=$\Sexpr{round(mixProp,2)}, we have that
$\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
Estimated thusly, the indirect effect is negative: \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
a standard error of \Sexpr{round(effects$tot_eff$std.error*(mixProp),2)}.
Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
With this important caveat, this analysis suggests that the
wide availability of a hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
treatment group:
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)Y(0,0)
\end{equation*}
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
\end{equation}
where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
\Sexpr{citopm(dir1.1ci)}.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
\Sexpr{citopm(dir2.1ci)}.

@@

\begin{table}
\centering
\begin{tabular}{lll}
\hline
Effect&Estimate&CI\\
\hline
<<mediationTable,echo=FALSE>>=
cat(
'Avg. Indirect Effect ($\\Delta$)&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
Avg. Direct Effect ($\\Xi$)&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
Avg. Direct Effect, covariate adjusted ($\\Xi$) &',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\\\\\n')
@
\hline
\end{tabular}
\caption{Estimated effects in mediation analysis. }
\label{tab:mediation}
\end{table}

@@

>>>>

The following section will provide background for the case-study,
describing the CTA1 curriculum, the RAND study, and our focus on
hints.
The next three sections will present an observational study,
which discards the control group altogether and estimates the effect
of hint requests on posttest scores;
causal mediation, which decomposes the overall treatment effect into a
component driven by hint requests and another driven by other
mechanisms; and principal stratification, which estimates separate
treatment effects for groups of students who would request hints
differently, if assigned to treatment.
As it turns out, these analyses appear to give contradictory
results---the observational study and mediation analysis estimate a
negative effect of requesting hints, while the principal
stratification analysis finds that students who request more hints
experience larger overall treatment effects.
Section \ref{sec:synthesis} will contrast these findings (including a
possible reconciliation), compare the three methods generally,
and discuss how they may be applied to data from other educational
technology RCTs.
Section \ref{sec:conclusion} will conclude.

@@

\subsection{Estimating the Effects of $H$ on Posttest Scores}\label{sec:obsEst}

@@

\section{Mediation Analysis}\label{sec:mediation}
The goal of causal mediation analysis is to decompose a treatment
effect into an ``indirect'' or ``mediated''
effect and a ``direct'' effect.
The indirect effect captures the component of the effect that is due
to the mediator: the treatment affected the outcome by affecting the
mediator, which itself, in turn, affected the outcome.
Assignment to the treatment condition allowed students to request many
hints during their work---how did the wide availability of hints affect their posttest scores?
The direct effect is the component of the overall effect that operates via other mechanisms.
For instance, how would assignment to CTA1 affect posttest scores were
hints more limited?

@@

It is worth noting here that conducting mediation analysis requires
measuring or positing a value of the mediator in the control arm of
the study.
We know that CTA1 hints were not available to the control group, so we
set $H=0$.
However, other facets of log data may not be defined, or coherent, in
the control group; for instance, students' proclivity to master the
skills of one CTA1 section before moving on to the next \citep{aoas}
or teachers' practice of reassigning students to new CTA1 sections
\citep{reassignmentEffect}.
In cases such as these, the mediation framework may not apply.

The following subsection reviews the formal definitions of direct and indirect effects.
Subsection \ref{sec:medID} discusses identification, and
\ref{sec:indirectEst} and \ref{sec:directEst} discuss estimation of
indirect and direct effects, respectively.

\subsection{Review: Defining Direct and Indirect Effect}
Express a subject's potential outcomes as a function
of two variables, treatment assignment $Z$ and hint requests $H$:
$Y(Z,H)$.
Now, if $Z$ and $H$ are both binary, as in our example, subjects
each have four potential outcomes:
\begin{equation*}
Y(0,0);\;Y(1,0);\;Y(0,1);\;Y(1,1)
\end{equation*}
representing the outcomes they would exhibit were they assigned to
control and requested few hints, assigned to
treatment and requested few hints, assigned to control and requested
many hints, or assigned to treatment and requested many hints, respectively.
In the CTA1 trial, however, the potential outcome $Y(0,1)$ is
meaningless: students in the control group had no access to CTA1, and
therefore could not request hints.
This will be an important factor going forward.

@@

Combining potential values for $Y$ and for $H$ yields the fundamental
building blocks of causal mediation analysis
\citep[e.g.][]{vanderweele2015explanation}.
Let $Y(z,H(z'))$
represent the outcome that a subject would express given treatment
assignment $z$, but the hint behavior that he or she would have
expressed under assignment $z'$.
When $z=z'$, the original two potential outcomes for $Y$ emerge:
$Y(1,H(1))=Y(Z=1)$ is the outcome exhibited under the treatment, when $H$
takes its potential treatment value; similarly, $Y(0,H(0))=Y(Z=0)$. On
the other hand, $Y(1,H(0))$ and $Y(0,H(1))$ are strictly
counterfactual. $Y(1,H(0))$ gives the outcome value
that would result for a subject assigned to the treatment condition
but who nevertheless requested hints as he or she would have under
control. $Y(0,H(1))$ gives the potential outcome for a subject
assigned to control but who nevertheless requested hints as he or she
would have under treatment.
This last quantity is problematic in our case, since $H(1)$ can take the value 1,
and as we have seen $Y(0,1)$ is undefined.

This framework facilitates precise definitions of direct and indirect
effects; in fact, there are two versions of each \cite[e.g.][]{imai2011unpacking}:
\begin{align*}
\delta(1)&=Y(1,H(1))-Y(1,H(0))\\
\delta(0)&=Y(0,H(1))-Y(0,H(0))\\
\end{align*}
are the indirect effects, contrasting potential outcomes when $H$
varies as it would with varying treatment assignment, but holding the
assignment itself constant at either 1 or 0.
\begin{align*}
\xi(1)&=Y(1,H(1))-Y(0,H(1))\\
\xi(0)&=Y(1,H(0))-Y(0,H(0))\\
\end{align*}
represent the direct effects: holding the value of $H$ constant,
either at its potential value under treatment or control, while
varying treatment assignment.
The total treatment effect, $Y(1,H(1))-Y(0,H(0))$, can then be
decomposed in two ways: as $\delta(1)+\xi(0)$ or as
$\delta(0)+\xi(1)$.
These decompositions can potentially reveal the role hints play in the
CTA1 treatment effect: $\delta(1)$ gives the extent to which
assignment to CTA1 affects a student's posttest score by (possibly)
causing him or her to request many hints.
$\xi(0)$ gives the effect of assignment to CTA1 if hints were,
perhaps counterfactually, held at their control level, $H=0$.

Rather than attempt to estimate individual direct and indirect effects
$\xi_i$ and $\delta_i$, our goal will be to estimate their means,
$\EE \xi$ and $\EE \delta$.

\subsection{Identification of Direct and Indirect
  Effects}\label{sec:medID}

@@

Dichotomizing hint requests into $H$ provides a
way forward.
Since the experimental design ensures that $H(0)=0$, $Y(1,H(0))\equiv
Y(1,0)$; this potential outcome is observed for subjects in
the treatment arm with $H=0$.
As above, $Y(0,H(0))=Y(0,0)$ and $Y(1,H(1))$ are the observed outcomes
for subjects in the control and treatment arms, respectively.
Therefore, we observe each of the potential outcomes in the definitions of
$\delta(1)$ and $\xi(0)$ for at least some non-trivial
portion of subjects in the study sample.

Of course, $H=0$ means something different in the two treatment arms:
in the treatment group $H=0$ typically means requesting
\emph{some} hints, but not many, whereas subjects in the control group
request exactly zero hints.
One way to make sense of mediational estimands involving $H$ is to
imagine an experiment in which students are assigned to different
values of $H$.
\Sexpr{round((1-mixProp)*100)}\% of students in the treatment group
are randomly assigned to $H=1$ and \Sexpr{round((mixProp)*100)}\% to
$H=0$.
This assignment constrains $\bar{h}$, the proportion of problems on which they
may request hints, so that if $H_i=0$,
$\bar{h}_i$ is constrained to be less than
\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))},
but if $H_i=1$ $\bar{h}_i$ must be greater than or equal to
\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
In contrast, all students in the control group are assigned $H=0$ and $\bar{h}=0$.

Average potential outcomes $\EE Y(1,H(1))$ and $\EE Y(0,H(0))$ are
identified due to the randomization of $Z$, but estimating $\EE
Y(1,H(0))$ requires imputing the
potential outcome $Y(1,0)$ for treated subjects with $H_i=1$.
However, it turns out that the matching estimators from
\ref{sec:observational} already solved this problem:
under matched ignorability, within matched sets, $Y(1,0)$
is independent of $H$, implying that $\EE[Y(1,0)|M,H=1]=\EE[Y(1,0)|M,H=0]$.
That suggests imputations
\begin{equation}\label{eq:estY10}
\hat{Y}_i(1,0)=\begin{cases}
 Y_i & H_i=0\\
\bar{Y}_{M=m_i,H=0} & H_i=1
\end{cases}
\end{equation}
That is, when $Z_i=1$ and $H_i=0$, no imputation is necessary, since
$Y_i(1,0)=Y_i$, the observed outcome. When $Z_i=1$ and $H_i=1$,
impute $Y(1,0)$ as the average of the observed outcomes from the $H=0$ subjects
matched to $i$.
Matched ignorability implies that the imputations will be unbiased.

\subsection{Estimating Log Data Indirect Effects}\label{sec:indirectEst}

Since $\EE \delta(1)$ involves only $Z=1$ potential outcomes, and
$H(0)=0$ is known based on the study design, $\EE\delta(1)$ may be
estimated using data from the treatment group only.

Using the imputations from \eqref{eq:estY10}, we estimate
$\EE\delta(1)$ as
\begin{equation*}
\widehat{\EE\delta(1)}=\widehat{\EE Y(Z=1)}-\widehat{\EE
  Y(1,0)}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,0)
\end{equation*}
where $n_T=\sum_i Z_i$, the number of subjects in the treatment
group.
Since $Y_i=\hat{Y}_i(1,0)$ whenever $H_i=0$, the summand is non-zero
only when $H_i=1$.
Further, note that when multiple $H=1$ subjects share a
matched set, they also share imputations $\hat{Y}(1,0)$.
Re-writing $\widehat{\EE\delta(1)}$ in terms of matched sets $M=m$,
\begin{align}
\widehat{\EE\delta(1)}&=\frac{1}{n_T}\displaystyle\sum_m\displaystyle\sum_{i:H_i=1,M_i=m}
                        Y_i-\bar{Y}_{M=m_i,H=0} \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_m
  n_{1m}(\bar{Y}_{M=m_i,H=1}-\bar{Y}_{m=m_i,H=0}) \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_mn_{1m}\hat{\tau}_m=\frac{\sum_iH_i}{n_T}\widehat{TOT} \label{eq:indTot1}
\end{align}
where $n_{1m}$ and $\hat{\tau}_m$ were defined in Section
\ref{sec:obsEst} as the number of $H=1$ subjects in match $m$ and the
difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
in match $m$, respectively, and $\widehat{TOT}$ is the TOT estimate,
\eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
The upshot of \eqref{eq:indTot1} is that under matched ignorability,
$\EE\delta(0)$ can be estimated as the TOT times the proportion of the
treatment group with $H=1$.

Actually, this relationship between $\EE\delta(0)$ and the TOT holds
in the population, regardless of how the TOT is estimated.
This is because when $H_i=0$, then $H_i(1)=H_i(0)=0$, so the individual direct effect
$\delta_i(1)=0$, and when $H_i=1$,
 $\delta_i(1)=Y_i(H=1)-Y_i(H=0)$, the effect of $H$ on $Y$.
Therefore,
\begin{equation*}
\EE \delta(1)=\EE\EE[\delta(1)|H]=\EE[\delta(1)|H=1]Pr(H=1)=TOTPr(H=1)
\end{equation*}

In the hint study, the definition of $H$ implies that
$Pr(H=1)=\sum_iH_i/n_T=$\Sexpr{round(mixProp,2)}.
Therefore, our estimate of the average indirect effect is
$\widehat{\EE \delta(1)}=$\Sexpr{round(mixProp,2)}$\widehat{TOT}$.
In Table \ref{tab:matchResults}, we estimated the TOT as
\Sexpr{round(effects[['tot_eff']][['coefficients']],2)} with a
standard error of \Sexpr{round(effects$tot_eff$std.error,2)}.
That implies an average indirect effect of \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
a standard error of \Sexpr{round(effects$tot_eff$std.error*(mixProp),2)}.

It is worth repeating that these estimates assume matched ignorabilty---no confounding between
$H$ and $Y$ within matched sets.
With this important caveat, this analysis suggests that the
wide availability of a hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate average direct effects,
$\EE[\xi(0)]=\EE[Y(1,H(0))-Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
  is equivalent to the ``controlled direct effect,'' $CDE(0)$
  e.g. \citet[][p. 57]{vanderweele2015explanation}.}
In principal, this could be estimated as the difference in sample
means of $\hat{Y}(1,0)$ in the treatment group and observed $Y=Y(0,0)$
in the control group.
In general, if we let
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)Y(0,0)
\end{equation*}
then we can estimate $\EE[\xi(0)]$ as the ATE of treatment assignment
$Z$ on $\tilde{Y}$.
Just like estimates of the ATE of $Z$ on $Y$, this estimate should
account for the experimental design: a paired, group-randomized trial
with students clustered in schools and schools randomized within
pairs.

For the sake of simplicity, we will use an OLS regression estimator,
with cluster-robust standard errors \citep{pustejovsky2018small},
clustered at the school level.
(Since in Section \ref{sec:observational} subjects were matched
within schools, school-level cluster-robust standard errors also account for the
dependence of $\hat{Y}(1,0)$ between matched treated students.)
Specifically, we regress:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{p[i]}+\widehat{\EE[\xi(0)]}Z_i+\epsilon_i
\end{equation}
where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
of subject $i$'s school and $\epsilon_i$ is a regression error.
Using the \texttt{clubSandwich} package in \texttt{R}
\citep{clubsandwich} to estimate standard errors, and 95\% confidence intervals, we estimated a direct
effect of
\Sexpr{citopm(dir1.1ci)}.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
\Sexpr{citopm(dir2.1ci)}.

@@

\begin{table}
\centering
\begin{tabular}{lll}
\hline
Effect&Estimate&CI\\
\hline
<<mediationTable,echo=FALSE>>=
cat(
'Avg. Indirect Effect ($\\EE \\delta(1)$)&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
Avg. Direct Effect ($\\EE\\xi$)&',round(dir1.1t[1,1],2),'&',printci(dir1.1ci),'\\\\
Avg. Direct Effect, covariate adjusted ($\\EE\\xi$) &',round(dir2.1t[1,1],2),'&',printci(dir2.1ci),'\\\\\n')
@
\hline
\end{tabular}
\caption{Estimated effects in mediation analysis. }
\label{tab:mediation}
\end{table}

@@

When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
$Y_i(1,0)$ is estimated with the average of their observed outcomes.

Now, for each member $i$ of the treatment group we have observed
$Y_i(1,H(1))=Y_i$, and estimated or inferred
$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
Then, we can estimate $\EE \delta(1)$, the average indirect effect in the
treatment group, by averaging over the treatment group:
\begin{equation*}
 \hat{\EE \delta(1)}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
\end{equation*}

This estimate is a close cousin of the
